---
title: "Homework 3"
format: pdf
author: "Elise LaFreniere"
date: 05-28-2026
execute: 
  message: false
  warning: false 
---

# Part 1. Set up tasks

```{r}
#| label: packages-and-data 
# reading in packages 
library(tidyverse)
library(janitor)
library(here)

# reading in kelp data 
kelp <- read.csv(
  here("data", "temp-kelp.csv")
)

# reading in personal data finalized
```

# Part 2. Problems 

## Problem 1. Giant kelp fronds 
Climate change threatens all ecosystems, including giant kelp (*Macrocystis pyrifera*) forests.

You are interested in the potential relationship between ocean temperature (measured in $^{\circ}$C) and giant kelp frond elongation rate (also known as growth rate, measured in cm day^-1^).

You conduct a field study in which you track giant kelp individuals growing in different temperatures (on average) and measure their elongation rates. Admittedly, this is not a perfect study, but you do the best with what you have!

#### a. An appropriate test

The appropriate tests to determine the strength of the relationship between ocean temperature and giant kelp frond elongation rate are Pearson's correlation ($r$) and Spearman's rank correlation ($\rho$) test. Pearson's correlation ($r$) evaluates a linear relationship between the continuous variables and assumes normality of the data, whereas Spearman's rank correlation ($\rho$) assesses the monotonic relationship by ranking the data points. Since biological growth rates like kelp elongation often do not follow a typical linear pattern, Spearman's rank correlation ($\rho$) may provide a more accurate measure of the relationship strength. 

#### b. Create a visualization 

```{r}
# base layer: ggplot
ggplot(kelp, # from data source
       mapping = aes(
         x = temp_c, # x axis temp
         y = kelp_elong # y axis elongation
       )) +
  # first layer : data points showing individual observation 
  geom_point(size = 2, 
             alpha = 0.7,
             color = "palegreen4") + # adding custom color 
  # adding a linear trend line
  # adding ribbon to represent confidence interval 
  geom_smooth(method = "lm",
              color = "sienna",
              fill = "gray",
              alpha = 0.3) +
  # renaming axes, adding a title 
  labs(
    x = "Temperature (°C)",
    y = "Kelp Elongation (cm/day)", 
    title = "Giant kelp frond elongation rate in
response to ocean temperature"
  ) +
  # adding theme 
  theme_minimal()
```

Ocean temperature increase leads to an approximate decrease in kelp elongation. 

#### c. Check your assumptions and run your test

##### Checking assumptions for Pearson's correlation (r)

```{r}
#| fig-width: 9
#| fig-height: 8
# writing out model in code
kelp_model <- lm(
  kelp_elong ~ temp_c,
  data = kelp)
# looking at diagnostics 
par(mfrow = c(2,2)) # creating 2 x 2 grid
plot(kelp_model) # plotting diagnostics 
```

**Diagnostic Assessment**

- **Residuals vs Fitted & Scale-Location:** The residuals vs fitted and scale-location appear to be homoscedastic through the approximate match of the red line to the trend line and scatter of points in relation to the line, indicating constant variance. 
- **QQ-Residiuals:** The QQ residuals points approximately follow the reference line with slight deviation at the left and right tails, allowing the assumption of normality to be met. 
- **Residuals vs Leverage:** No outliers are past Cook's distance, indicating that outliers are not influencing the model coefficient estimates. 

Since the assumptions of normality and homoscedasticity are met, and no influential outliers are present, a linear model framework is appropriate for this data, validating the use of a Pearson's correlation test ($r$).

##### Running a Pearson's correlation test (r)

```{r}
# creating a new object to display correlation test results
kelp_correlation <- cor.test(kelp$temp_c,
                             kelp$kelp_elong, 
                             # pearson's correlation
                             method = "pearson")
# viewing correlation test results 
kelp_correlation
```

#### d. Results communication


