---
title: "Homework 3"
format: pdf
author: "Elise LaFreniere"
date: 05-28-2026
execute: 
  message: false
  warning: false 
---

# Part 1. Set up tasks

```{r}
#| label: packages-and-data 
# reading in packages 
library(tidyverse)
library(janitor)
library(here)

# reading in kelp data 
kelp <- read.csv(
  here("data", "temp-kelp.csv")
)

# reading in personal data finalized
```

# Part 2. Problems 

## Problem 1. Giant kelp fronds 
Climate change threatens all ecosystems, including giant kelp (*Macrocystis pyrifera*) forests.

You are interested in the potential relationship between ocean temperature (measured in $^{\circ}$C) and giant kelp frond elongation rate (also known as growth rate, measured in cm day^-1^).

You conduct a field study in which you track giant kelp individuals growing in different temperatures (on average) and measure their elongation rates. Admittedly, this is not a perfect study, but you do the best with what you have!

#### a. An appropriate test

The appropriate tests to determine the strength of the relationship between ocean temperature and giant kelp frond elongation rate are Pearson's correlation ($r$) and Spearman's rank correlation ($\rho$) test. Pearson's correlation ($r$) evaluates a linear relationship between the continuous variables and assumes normality of the data, whereas Spearman's rank correlation ($\rho$) assesses the monotonic relationship by ranking the data points. Since biological growth rates like kelp elongation often do not follow a typical linear pattern, Spearman's rank correlation ($\rho$) may provide a more accurate measure of the relationship strength. 

#### b. Create a visualization 

```{r}
# base layer: ggplot
ggplot(kelp, # from data source
       mapping = aes(
         x = temp_c, # x axis temp
         y = kelp_elong # y axis elongation
       )) +
  # first layer : data points showing individual observation 
  geom_point(size = 2, 
             alpha = 0.7,
             color = "palegreen4") + # adding custom color 
  # adding a linear trend line
  # adding ribbon to represent confidence interval 
  geom_smooth(method = "lm",
              color = "sienna",
              fill = "gray",
              alpha = 0.3) +
  # renaming axes, adding a title 
  labs(
    x = "Temperature (°C)",
    y = "Kelp Elongation (cm/day)", 
    title = "Giant kelp frond elongation rate in \n response to ocean temperature"
  ) +
  # adding theme 
  theme_minimal()
```







